\documentclass[12pt]{jhwhw}
\author{Ian Malerich}
\title{Com S 331: Homework 11}
\usepackage{amssymb, amsfonts, amsthm, mathtools, graphicx, breqn}
\usepackage{subfig, float, scrextend, setspace, soul}
\usepackage{tikz, subcaption, caption, graphviz, color}
\usetikzlibrary{arrows.meta, automata, positioning}

\newcommand{\hlred}[1]{{\sethlcolor{pink}\hl{#1}}}
\newcommand{\hlgreen}[1]{{\sethlcolor{lime}\hl{#1}}}

\onehalfspacing
\begin{document}
\raggedright

%% Problem 1
\problem{}

	You are employed as a programmer, and you are asked to write the given program.
	For each case, write ``Y" or ``N" if the program can or cannot 
	be written. Do not concern yourself with memory limitations, that is, assume
	that the computer used to run your program has a large, effectively unbounded,
	amount of memory.

\solution

\part %% (a)
	\begin{center}
	\textit{
		receives in input a generic C program $x$, 
		and counts the number of statements in $x$.
	}
	\end{center}
	Y

\part %% (b)
	\begin{center}
	\textit{
		receives in input a generic C program $x$,
		and an input string $w$, and counts the number statements executed
		at least once when $x$ runs on input $w$.
	}
	\end{center}
	N

\part %% (c)
	\begin{center}
	\textit{
		receives in input a generic C program $x$ and an input string  $w$, and 
		counts the number of statements never executed when $x$ runs on
		input $w$.
	}
	\end{center}
	N

\part %% (d)
	\begin{center}
	\textit{
		receives an input a generic C program $x$ and decides whether $x$ is
		syntactically correct.
	}
	\end{center}
	Y

\part %% (e)
	\begin{center}
	\textit{
		receives in input two natural numbers and computes a specific function
		$f: \mathbb{N}^2 \rightarrow \mathbb{N}$.
	}
	\end{center}
	N

\part %% (f)
	\begin{center}
	\textit{
		receives in input a generic arithmetic expression $e$ composed of integers
		and the four arithmetic operators, and computes its value.
	}
	\end{center}
	Y

\part %% (g)
	\begin{center}
	\textit{
		halts on the empty string
	}
	\end{center}
	Y

\part %% (h)
	\begin{center}
	\textit{
		receives in input a generic C program $x$ and decides whether $x$ halts
		only on the empty string.
	}
	\end{center}
	N

\part %% (i)
	\begin{center}
	\textit{
		receives in input two generic regular expressions and decides whether
		they are equivalent.
	}
	\end{center}
	Y

\part %% (j)
	\begin{center}
	\textit{
		receives in input a generic C program $x$ and the name of one of its functions, f,
		and decides whether $x$ can ever call $f$.
	}
	\end{center}
	N

\part %% (k)
	\begin{center}
	\textit{
		receives in input a generic C program $x$, an input string $w$, and the name
		of one of its functions, $f$, and decides whether $x$ calls $f$ when
		running on input $w$.
	}
	\end{center}
	N

\part %% (l)
	\begin{center}
	\textit{
		receives in input two generic C program $x_1$ and $x_2$ and an input
		string $w$, and decides whether $x_1$ and $x_2$ produce the same output
		when running on input $w$.
	}
	\end{center}
	N

\part %% (m)
	\begin{center}
	\textit{
		receives in input two generic C programs $x_1$ and $x_2$ and decides
		whether $x_1$ and $x_2$ produce the same output when running on every
		possible input.
	}
	\end{center}
	N

\part %% (n)
	\begin{center}
	\textit{
		receives in input two generic C programs $x_1$ and $x_2$, and decides
		whether $x_1$ and $x_2$ produce the same output when running on at least one
		input.
	}
	\end{center}
	N

\part %% (o)
	\begin{center}
	\textit{
		receives in input a generic C program $x$, an input string $w$, and a natural
		number $n$, and decides whether $x$ uses less than $n$ bytes of memory when 
		running on $w$.
	}
	\end{center}
	Y

\part %% (p)
	\begin{center}
	\textit{
		receives in input a generic C program $x$, an input string $w$, and 
		decides whether there is an $n\in \mathbb{N}$ such that $x$ uses less
		than $n$ bytes of memory running on input $w$.
	}
	\end{center}
	N

%% Problem 2
\problem{}

	Use reduction to prove that the language
	$$
		L = \{p(M)p(w) :\text{ the TM $M$ never enters its initial state again when
			running on $w$}\}
	$$
	is undecidable.

\solution

	\begin{proof} \textit{via contradiction} \\
		Assume that $L$ is decidable $\Rightarrow \exists$ a Turing Machine $K$ which generates $L$.
		Define a new Turing machine $K'$ which takes as input a Turing machine $M$ and string $w$
		encoded as $p(M)p(w)$. $K'$ will perform a series of modifications to the input machine $M$,
		we will call this modified machine $M'$.
		Denote the initial state of $M$ as $q_0$, $K'$ will create a new state $q_0'$, set it as the
		new initial state for $M'$ and add an $\epsilon$-transition from $q_0'\rightarrow q_0$. Thus
		$M$ halts $\Leftrightarrow M'$ halts. Note that $M'$ will never enter state $q_0'$ again 
		when running.
		\bigbreak
		Next take the accepting and rejecting states of $M$, call them $a$ and $r$, these will no longer
		be the halting states of $M'$, but otherwise exist the same. Instead introduce new states $a'$ and $r'$
		which are the respective accepting and rejecting states, however these include no edges. Next add
		$\epsilon$-transitions from $a$ and $r$ to $q_0'$. 
		Then $M$ halts $\Leftrightarrow$ $M$ enters either
		state $a$ or $r$ $\Leftrightarrow$ $M'$ enters state $q_0'$ again$^{\textcolor{red}{[1]}}$. 
		Further, if $M$ does not halt then
		$M$ does not enter states $a$ or $r$. We know the only transitions into $q_0'$ are through $a$ and $r$, we
		therefore can conclude that $M'$ does not enter state $q_0'$ again$^{\textcolor{red}{[2]}}$.
		\bigbreak
		Now that our machine $K'$ has set up this machine $M'$ based on its input, $K'$ will then run
		the machine $K$ recognizing $L$ on $M'$ and output the result. 
		Note that $M'$ is guaranteed to not halt, but that is not
		important or relevant to machine $K.$ Construction of our $M'$ is easily computable, further $K$ is guaranteed
		to return a result by assumption. Thus $K'$ is decidable.
		\bigbreak
		If $K'$ accepts then $q_0'$ is entered again $\Rightarrow$ $M$ halts$^{\textcolor{red}{[1]}}$. \\
		If $K'$ rejects then $q_0'$ is not entered again $\Rightarrow$ $M$ hangs$^{\textcolor{red}{[2]}}$. \\
		\bigbreak
		Thus $K'$ solves the halting problem $\Rightarrow$ $K'$ is undecidable. This contradicts
		our statement that $K'$ is decidable. Therefore our assumption is false and we conclude
		that $L$ is undecidable.
	\end{proof}

\end{document}
