\documentclass[12pt]{jhwhw}
\author{Ian Malerich}
\title{Com S 331: Homework 10}
\usepackage{amssymb, amsfonts, amsthm, mathtools, graphicx, breqn}
\usepackage{subfig, float, scrextend, setspace, soul}
\usepackage{tikz, subcaption, caption, graphviz}
\usetikzlibrary{arrows.meta, automata, positioning}

\onehalfspacing
\begin{document}
\raggedright

%% Problem 1
\problem{}

	Prove that the class of Turing-acceptable languages is closed under union, intersection, and reversal.
	For each property, give a detailed \underline{sketch} of the proof, by saying how you would
	build a Turing machine that accepts the resulting language, given the Turing machine(s) that
	accept the original language(s).

\solution

	\textbf{Union} \\
	\begin{proof}
		Let $L_1, L_2$ be Turing-acceptable languages, then define their corresponding 
		Turing machines as $M_1, M_2$ respectively.

		We will then define a new machine $M_U$ which operates on the union of
		$Q,\Sigma$ for $M_1, M_2$. We will distinguish each state by the source machine,
		thus $Q_1 \cap Q_2 = \emptyset$. The initial state of $M_U$ will be the initial
		state of $M_1$.
		
		On the input of a string $w$, $M_U$ will behave exactly as $M_1$. If 
		we reach a halting and accepting state of $M_1$, then $M_U$ will also halt and accept.
		Otherwise, rather than halting and accepting, $M_U$ will then run $M_2$ on $w$, 
		once $M_2$ halts, $M_U$ accepts if $M_2$ does, and rejects otherwise.
		Note that $M_I$ will halt provided that both $M_1$ and $M_2$ each halt.

		Now assume $w\in L_1\cup L_2$, then either $w\in L_1$ or $w\in L_2$, if $w\in L_1$
		$w$ will be accepted by $M_U$ after the initial run of $M_1$ (as $M_1$ accepts $w$).
		Otherwise if $w\in L_2$, then after $M_U$ runs $M_1$ where $M_1$ rejects $w$,
		$M_U$ will continue with $M_2$ which accepts $w$ and thus so does $M_U$.
		If $w\not\in L_1\cup L_2$ then $w\notin L_1$ thus $M_U$ continues onto $M_2$,
		but as $w\not\in L_2$, $L_2$ will reject $w$ and thus so will $M_U$.

	\end{proof}

	\bigbreak
	\textbf{Intersection} \\
	\begin{proof}
		Let $L_1, L_2$ be Turing-acceptable languages, then define their corresponding 
		Turing machines as $M_1, M_2$ respectively.

		We will then define a new machine $M_I$ which operates on the union of
		$Q,\Sigma$ for $M_1, M_2$. We will distinguish each state by the source machine,
		thus $Q_1 \cap Q_2 = \emptyset$. The initial state of $M_U$ will be the initial
		state of $M_1$.

		On the input of a string $w$, $M_I$ will behave exactly as $M_1$. If we reach
		a halting and rejecting state of $M_1$, then $M_I$ will halt and reject, if $M_1$
		halts and accepts, $M_I$ will continue on and run $M_2$ on $w$. If $M_2$ accepts
		$w$, then so will $M_I$ otherwise $w$ is rejected by $M_I$.
		Note that $M_I$ will halt provided that both $M_1$ and $M_2$ each halt.

		Now assume $w\in L_1\cap L_2$, then $w\in L_1$ and $w\in L_2$, because of this,
		$w$ will run through both stages of $M_I$ and pass each, thus $M_I$ clearly accepts $w$.
		However, if $w\not\in L_1\cap L_2$, then $w\not\in L_1$ (in which case
		$w$ is rejected in the first stage of $M_I$) or $w\not\in L_2$ (in which case
		$w$ is rejected by the second stage of $M_I$). In either case, $w$ will be rejected
		by $M_I$, therefore $M_I$ represents $L_1\cap L_2$.
	\end{proof}

	\bigbreak
	\textbf{Reversal} \\
	\begin{proof}
		Let $L$ be a Turing-acceptable language, then define a corresponding Turing machine $M$
		which. Define a new Turing machine $M_R$ using the same alphabet, and set of states as $M$.
		However let the initial state of $M_R$ be the final accepting state of $M$, similarly, let
		the final state of $M_R$ be the initial state of $M$.

		We can then run $M$ backwards on $w$ (note we are starting on the final state), if we
		can then reach the initial state of $M$ we have successfully read the reverse of a string
		$w\in L$, thus the initial state of $M$ will be the final accepting state of $M_R$. If 
		running $M$ backwards halts but does not accept, then $M_R$ should reject the input.
		In this sense, the $\delta$ function of $M_R$ is the inverse of the $\delta$ function for $M$.

		To show this accepts reversal, assume $w\in L$ then $M$ accepts $w$ and $w^R\in L^R$.
		Running $M_R$ on $w^R$ is equivalent to running $w$ on $M$, which accepts $w$ thus $M_R$
		accepts $w^R$. If $w\not\in L$ then $M$ does not accept $w$, thus $w^R\not\in L^R$.
		Running $M^R$ on $w^R$ is again equivalent to running $w$ on $M$, however as $w\not\in L$,
		we know that $M$ will not reach an accepting state, therefore running $\delta$ backwards from
		the accepting state cannot possibly reach the initial state (the final accepting state of $M^R$)
		therefore $M^R$ will not accept $w^R$.
	\end{proof}

%% Problem 2
\problem{}

	Prove or disprove that the set of Turing-acceptable languages is closed under concatenation.

\solution

\begin{proof}

	Let $L_1, L_2$ be Turing-acceptable languages, then define their corresponding 
	Turing machines as $M_1, M_2$ respectively.

	We will then define a new machine $M_C$ which operates on the union of
	$Q,\Sigma$ for $M_1, M_2$. We will distinguish each state by the source machine,
	thus $Q_1 \cap Q_2 = \emptyset$. The initial state of $M_U$ will be the initial
	state of $M_1$. We will also include a set of 'marked' characters in our alphabet,
	which correspond each possible state of $Q_1\times \Sigma_1$ (note this additional set of 
	characters is finite), these are unique from other characters in our alphabet.

	Our new machine $M_C$ will first run $M_1$ until it reaches any arbitrary accepting state,
	note $M_1$ need not have halted. $M_C$ then marks on the tape with one of the 'marked' symbols
	which corresponds to the state of $M_1$ and the current character at that point. We know
	the string up to this marked point is accepted by $M_1$ thus we then run $M_2$ on the remainder
	of the string, if $M_2$ accepts the remainder, then we halt and accept the string. If $M_2$ does not
	accept the remainder, we enter a new set of states which take us back to the marked location,
	rewrites the original character (remember that it is encoded into our marked character), restores
	the state of $M_1$ and repeats the process until the next accepting state of $M_1$ is reached.
	If we continue and run out of characters in the string for which $M_1$ could accept, we halt
	and reject the string.
	
	By performing this process, $M_C$ divides an input string $w$ into two substrings $w=xy$, where $x$
	is accepted by $M_1$, the machine then checks if $y$ is accepted by $M_2$ if so then $w$ must
	be in the concatenation by definition. If not, we grow $x$ until it is again accepted by $M_1$, 
	and perform our check once again. Clearly then if $w$ is in the concatenation of $L_1, L_2$ $M_C$
	will accept it by testing every possible combination until a division is found that works.
	If $w$ is not in the concatenation of $L_1,L_2$, then we must eventually reach a point where no
	more strings can be accepted by $M_1$ (we are trying to feed a string longer than our initial string
	into $M_1$) at this point our machine will halt and reject the string.

\end{proof}

%% Problem 3
\problem{}

	Consider a new type of \textit{deterministic} machine, having one read-only input tape
	and two stacks. The tape is read-only, it cannot be written, but the head can move left, right,
	or do nothing. Each stack operates (independently of the other) as in a deterministic pushdown
	automaton.
	$$
		M = (K,\Sigma, \Gamma_1, \Gamma_2, z_1, z_2, \delta, s)
	$$
	where $K$ is a finite set of states, $\Sigma$ is a finite input alphabet,
	$\Gamma_1$ and $\Gamma_2$ are two finite stack alphabets, $z_1\in \Gamma_1$ and $z_2\in \Gamma_2$
	are the initial symbols for the two stacks, $s\in K$ is the initial state. $h$ is a special
	halting state not in $K$, just like a Turing machine.

	\begin{enumerate}
		\item Give an appropriate definition for the transition function $\delta$, for a 
			configuration of this machine, for the ``yields in one step" operator,
			and for the language accepted by this machine.
		\item These machines can accept the same languages as a class of automata you already know:
			deterministic pushdown automata, pushdown automata, or Turing machines? Prove
			your answer formally.
	\end{enumerate}

\solution

	\part

	The $\delta$ function on input will know it's current state, the current character under it's head,
	as well as the character at the top of each of it's stacks. On output it will produce
	either a new state or halting state, the head can move left/right/stay, and it can push/pop a character
	onto/off either of it's stacks. Note this transition function is basically the transition function
	of NPDA glued together with the transition function of Turing machines, with the exception
	of the absence of the output character which the Turing machine writes back to the stack.

	$$
		\delta: (K\times\Sigma\times\Gamma_1\times\Gamma_2) \rightarrow 
			((K\cup \{h\})\times \{L,R,\epsilon\}\times\Gamma_1^*\times\Gamma_2^*)
	$$

	\part
	This machine is equivalent to a Turing machine.
	\begin{proof}
		Let $T$ be a Turing machine which represents some language $L$.
		We will construct an equivalent machine $M$ using our new type of machine to $T$.
		Let $\Sigma_T$ be the alphabet used by the Turing machine.
		Then our machine $M$ will have $\Sigma=\Sigma_T,\Gamma_1=\Sigma_T,\Gamma_2=\Sigma_T$, that is
		both our tape and both stacks have the same alphabet as the Turing machine tape.
		Our set of states will be equivalent to those of the Turing machine, with the exception
		of two additional setup states which I will denote $S_M$ and $S_C$. $S_M$ will serve
		as the initial state of $M$.
		\bigbreak
		\textbf{SETUP} The tape for machine $M$ is initialized to the same state as the Turing machine $T$.
		When in state $S_M$, the delta function will move the head to the left (remember
		that the machine starts at the end of the input) and will repeat this process until
		it reaches either the end of the tape or the first input character (if this is an infinite tape,
		we could overstep, read nothing, then step back). The machine then enters state $S_C$, in which
		it will read a single character, push that character on to it's first stack, then move one character
		to the right. Once it reads an empty character (has reached the end of its input) it enters 
		the initial state of the Turing machine.
		\bigbreak
		\textbf{CLAIM} Stack 1 represents characters to the left of the head in the Turing machine,
		and Stack 2 represents characters to the right (in the initial state of the Turing machine).
		This is clear to see, since we read from left to right while pushing onto the stack, Stack 1
		must necessarily the characters to the right of the Turing machine head (with nearest being
		closer to the top of the stack), the Turing machine has nothing to its right therefore
		Stack 2 (which is empty save $z_2$) is equivalent. Note that the top of Stack 2 is considered
		by my convention to by the location of the head.

		For the following operations on Turing machine $T$, I will define an operation on $\delta$ which
		I claim will be equivalent. In all cases, machine $M$ enters it's equivalent state that $T$ does. 
		Note that for move operations if we have an empty stack (save $z_i$) we can ignore the pop
		and for what I describe as the ``same value" below, instead use the empty character for the
		corresponding push. Also note that once we have read the tape, because of the
		fact that our stack now represents our tape, we never have to move the actual head we 
		will instead simulate with the stacks as shown below. \\
		\bigbreak
		\textbf{$T$ WRITES TO HEAD} $M$ pops from Stack 2 and pushes the write value to Stack 2. \\
		\begin{addmargin}{1em}
			Note that Stack 1 does not change, the Turing machine does not move, therefore
			the left of the head does not change. The Turing machine has a new value at it's head.
			By convention, this is top of Stack 2. When the top is removed, the value overwritten
			is removed now the item to the right of the head is at the top of the stack. After
			writing the top of Stack 2 now points to the new written value the same at the 
			head of the Turing machine. Thus these two operations are equivalent.
		\end{addmargin}
		\textbf{$T$ MOVES LEFT} $M$ pops from Stack 1 and pushes the same value to Stack 2. \\
		\begin{addmargin}{1em}
			The Turing machine has one less character to the left of it, what was 2 characters away
			is now 1 character away. After popping from Stack 1, $M$ has one less character to its left
			and what was 2 characters away is now 1 character away. \\
			What was under the Turing machine head is now to its right, what was to the left of the Turing
			machine is now under the head. After pushing the head of Stack 1, what was to the left of
			$M$ is now top of Stack 2 (that is, underneath the head of $M$). Further what was at the top
			of stack 2 is now one character away (as we pushed a new value).
		\end{addmargin}
		\textbf{$T$ MOVES RIGHT} $M$ pops from Stack 2 and pushes the same value to Stack 1. \\
		\begin{addmargin}{1em}
			Same as above, one less character to the right in $T$, where popping from Stack 2
			corresponds in one less character to the right in $M$. Pushing the same value to Stack 1
			is equivalent to having the character under the head now to the left of $T$ after moving.
			Again, note that if Stack 2 only contains $z_2$ do not pop from Stack 2, instead push the 
			empty character onto Stack 1. By the above it is clear that this will be equivalent, as an
			empty stack implies that there was nothing (infinitely many empty characters) to the right
			of the head.
		\end{addmargin}
		\textbf{$T$ HALTS AND ACCEPTS/REJECTS} $M$ also halts and accepts/rejects (same as $T$)
		\begin{addmargin}{1em}
			$T$ and $M$ are in equivalent states, thus when $T$ halts, $M$ can therefore also
			halt with the same result as $T$.
		\end{addmargin}

		Therefore in all situations, the machine we have defined represents the tape of $T$ through 
		use of its two stacks. Further, after completing its setup, $M$ will always be 
		in an equivalent state to $T$. Therefore, given the same input both $T$ and $M$ will perform
		the same operations and produce the same result. With the only difference being
		that $M$ is using stacks to represent the tape (initialized from its actual tape) where as
		$T$ can perform all its operations directly on the actual tape.

	\end{proof}

\end{document}
